Modal Abstractions for Virtualizing Memory Addresses

### 2025-03-21 relwork.tex
inline tlb appendix mostly into related work, update related work

There has been relatively little prior work on formal verification of virtual memory.
Instead, most OS verification work has focused on minimizing reasoning about virtual memory management.
The original \textsc{Verisoft} project~\cite{alkassar2008verisoft,alkassar2010pervasive,alkassar2008formal,dalinger2005verification,hillebrand2005address,alkassar2008formal,starostin2010formal} 
relied on custom hardware which, among other things, always ran kernel code with virtual memory disabled, removing the circularity that is a key challenge of verifying 
VMM code for real hardware: at that point page tables become a basic partial map data structure to represent user program address translations,
with an idiosyncratic format. It turns out that subsequent OS verification work also treats page tables this way, but
unsoundly given that other projects target hardware that \emph{does} run the kernel with address translation: they \emph{trust}
that the particular page table manipulations do not, for example, unmap kernel code (which can crash the machine even if done ``temporarily'').
% \footnote{A bug
% one author has personally encountered in a research kernel, though not one discussed here.}).
This is true of \textsc{seL4}~\cite{Klein2009seL4,seL4TOCS,Sewell2013translation}, whose formal machine model omits address translation;
and \textsc{CertiKOS}~\cite{gu15,gu2016certikos,gu2018certikos,chen2016interrupts}, whose refinement proofs rely on
\textsc{CompCert}'s usermode-oriented memory model~\cite{leroy2008formal,leroy2009formally} which assumes
updates to one memory address are independent of updates to another ---
which is not true of page table updates.
 % which is not the case when a single page table write can make regions of memory inaccessible.
% \textsc{seL4}~\cite{Klein2009seL4,seL4TOCS,Sewell2013translation} is a formally verified L4 microkernel~\cite{Liedtke1995,Liedtke1996} 
% (and the first verified OS kernel to run on real-world hardware), verified with a mix of refinement proofs and program logic reasoning down to the assembly level.
% Because \textsc{seL4} is a microkernel, most VMM functionality actually lives in usermode and is unverified, and moreover, 
% their hardware model omits address translation entirely and the MMU entirely~\cite{Klein2009seL4,seL4TOCS}. 
% As a result, the limited page table management present in the microkernel treats page tables as idiosyncratic tree-maps as in
% \textsc{Verisoft}, despite actually running with address translation.
% This is partly mitigated by manually identifying some trusted invariants (e.g., that the address range designated for the kernel is appropriately mapped)
% and setting up the proof to ensure those invariants are maintained (i.e., as an extra proof obligation not required by their hardware model).

% \textsc{CertiKOS}~\cite{gu15,gu2016certikos,gu2018certikos,chen2016interrupts} is a microkernel intended for use as a hypervisor.
% The overall approach in that body of work is many layers of refinement proofs, using a
%  proliferation of layers with small differences to keep most individual refinements tractable. In keeping with precursor work 
% on the project from the same group~\cite{vaynberg2012compositional}, the purpose of some layers is to abstract away from 
% virtual memory (as early as possible).
% The papers on \textsc{CertiKOS} do not explicitly detail the VMM beyond highlighting its existence and referencing
% that it performs mapping operations for user code.
% The work is clear, however, that it fully trusts low-level assembly fragments such as the instruction sequence which actually
% switches address spaces, rather than verifying them.
% We can also infer other limitations from the project's pervasive reliance on a modified \textsc{CompCert},
% which fixes an abstract memory model aimed at \emph{user-mode} C programs~\cite{leroy2008formal,leroy2009formally}. 
% \textsc{CompCert}'s model of memory cannot handle non-local updates, so we can conclude \textsc{CertiKOS}'s proofs cannot
% address impact of virtual memory mapping changes on the kernel itself.
Other work on OS verification either never progressed far enough to address VMM verification 
(\textsc{Verisoft XT}~\cite{cohen2009vcc,cohen2010local,dahlweid2009vcc,cohen2013SOFSEM}), or uses memory-safe languages to enable safe co-habitation 
of a single address space by all processes (\textsc{Singularity}~\cite{Fahndrich2006language,Hunt2007singularity,Hunt2007sealing,Barnett2011specsharp}, \textsc{Verve}~\cite{Yang2010Verve},
and \textsc{Tock}~\cite{levy2017multiprogramming}). The latter ensures memory safety, but ignores other functional uses of virtual memory
hardware like swapping~\cite{Denning1970VM} or exploiting copy-on-write techniques for dynamic migration of virtual machines~\cite{clark2005live}.

\citet{kolanski08vstte} are the only other researchers to study VMM verification against a realistic hardware
model, where page table updates are performed through virtual memory accesses (later adding C-level type modeling~\cite{kolanski09tphols}).
As noted in Section \ref{sec:overly-restrictive}, Kolanski and Klein's virtual points-to is similar to that in Figure \ref{fig:strongvirtualpointsto},
with the attendant problems discussed earlier and lifted by our model. Their approach had modal elements,
but the work never addressed the issues of managing multiple address spaces, switching address spaces, or
invariants referring to other address spaces, which our explicitly modal approach and case studies do address.
They verified a fraction of an ARM version of our Figure \ref{fig:mapping_code}, which assumed they already had the virtual address
of the L1 entry (which they assumed existed), thereby skipping the physical-to-virtual conversion issues we treat in Section \ref{sec:p2v}.
Due to lack of fractional permissions in their formalism, they incidentally pick up other limitations orthogonal to their
foundational focus: by requiring
a virtual points-to have \emph{full} ownership of the page table walk memory, they limit their system to having only as many
virtual points-to assertions as there are entries in the top-level table (4096) because they cannot share access to entries.
We inherit fractional permission support form Iris, and use it extensively (the overly-restrictive
Figure \ref{fig:strongvirtualpointsto} is already an improvement in this way).

@@

% The other major distinction is that Kolanski and Klein have no explicit accounting for other address spaces.
% Their logic does not deal with change of address space, and has no way to assert that certain facts hold
% in another address space.
% They verify only one address space manipulation: mapping a single unmapped page into the current address space (in both papers).
% We verify this, as well as a change-of-address-space, which requires us to introduce assertions for talking
% about other address spaces (we must know, for example, that the precondition of the code after the change must be true
% in the \emph{other} address space), and to deal with the fact that the standard frame rule
% for separation logic is unsound in the presence of address space changes and address-space-contingent assertions.
% Our approach in this paper uses modalities to distinguish virtual-address-based assertions that hold only in specific 
% address spaces, making it possible to manipulate other address spaces, and equally critically, to \emph{change} address 
% spaces while reasoning about correctness.
The only other attempt to verify a change of address space was by \citet{syeda2018program,syeda2020formal}, whose focus was on TLB maintenance (which we
do not model). However, they verified only the 4 instructions to switch address spaces and update the TLB on an ARM processor, in isolation
(i.e., not the full context switch including changing stacks with address spaces).
The specification they proved for those instructions did not address program invariants that may be valid in one address space and not the other,
so is not flexible enough to extend directly to a full context switching primitive as in Figure \ref{fig:swtch}.
Of course as we do not address TLB management (see Appendix \ref{apdx:tlb}), the right eventual solution would be to combine our work
and theirs, which we leave to future work.

\ifPLDI
\else
Unlike our work, Kolanski and Klein prove very useful embedding theorems stating that code that does not modify page table 
entries can be verified in a VM-ignorant program logic, and that proofs in that logic can be embedded into the VM-aware logic 
(essentially by interpreting ``normal'' points-to relations as virtual points-to facts). While we have not proven such a result,
an analagous result {should} hold of our work: consider that the doubles for the \texttt{mov} instructions
that access memory behave just as one would expect for a VM-ignorant logic~\cite{Chlipala2013Bedrock}.
With our general approach to virtual points-to assertions being inspired by Kolanski and Klein, \emph{both}
 our approach and theirs could in principle be extended to account for pageable points-to assertions by adding additional 
disjunctions to an extended points-to definition; embedding ``regular'' separation logic into such a variant
is the appropriate next step to extend reasoning to usermode programs running with a kernel that may demand-page the program's
memory.
\fi

As noted earlier, our other-space modality derives from hybrid logic~\cite{areces2001hybrid,blackburn1995hybrid,gargov1993modal,goranko1996hierarchies},
where modalities are indexed by \emph{nominals} which are names for specific individual states in a Kripke model.
Little prior work combines these ideas with program logics: \citet{brotherston2014parametric} show that traditional
nominals extends the expressive power of separation logic, and \citet{gordon2019modal}
uses nominals to refer to states of other actors in an actor language.
% As noted throughout the paper, the inspiration for our other-space modality comes from hybrid logic~\cite{areces2001hybrid,blackburn1995hybrid,gargov1993modal,goranko1996hierarchies},
% where modalities are indexed by \emph{nominals} which are names for specific individual states in a Kripke model.
% We are aware of only two prior works combining hybrid logics with program logics specifically. 
% Brotherston and Villard~\cite{brotherston2014parametric} demonstrated that may properties true of various 
% separation logics are not definable in boolean \BI (\BBI), and showed that a hybrid extension \HyBBI allows
% most such properties to be defined (e.g., the fact that separating conjunction is cancellative is unprovable 
% in boolean \BI, but provable in \HyBBI). There, nominals named resources 
% (roughly, but not exactly, heap fragments). 
% Gordon~\cite{gordon2019modal} described a use of hybrid logic in the verification of actor programs, 
% where nominals named the local state of individual actors (with such assertions stabilized with a 
% rely/guarantee approach). 
Beyond these, there is limited work on the interaction of hybrid logic with substructural logics, in restricted forms
that do not affect expressivity~\cite{despeyroux2014hybrid,chaudhuri2019hybrid}.\looseness=-1
% Primarily there is a line of work on hybrid linear logic (\HyLL)~\cite{despeyroux2014hybrid}, 
% originally used as a way to more conveniently express aspects of transition systems in linear logic. 
% However, \HyLL's proof rules offer no non-trivial interactions with multiplicative connectives 
% (every \HyLL proof can in fact be embedded into regular linear logic~\cite{chaudhuri2019hybrid}, 
% unlike Brotherston and Villard's \HyBBI, which demonstrably increases expressive power over its base \BBI.

>>>>

We will discuss related work along two axes: work on OS verification specifically, and work on 
program logics with modalities.

\subsection{OS Kernel Verification}
There has been relatively little prior work on formal verification of virtual memory.
Instead, most OS verification work has focused on minimizing reasoning about virtual memory management.
The original \textsc{Verisoft} project~\cite{alkassar2008verisoft,alkassar2010pervasive,alkassar2008formal,dalinger2005verification,hillebrand2005address,alkassar2008formal,starostin2010formal} 
relied on custom hardware which, among other things, always ran kernel code with virtual memory disabled, removing the circularity that is a key challenge of verifying 
VMM code for real hardware: at that point page tables become a basic partial map data structure to represent user program address translations,
with an idiosyncratic format. It turns out that subsequent OS verification work also treats page tables this way, but
unsoundly given that other projects target hardware that \emph{does} run the kernel with address translation: they \emph{trust}
that the particular page table manipulations do not, for example, unmap kernel code (which can crash the machine even if done ``temporarily'').
% \footnote{A bug
% one author has personally encountered in a research kernel, though not one discussed here.}).
This is true of \textsc{seL4}~\cite{Klein2009seL4,seL4TOCS,Sewell2013translation}, whose formal machine model omits address translation;
and \textsc{CertiKOS}~\cite{gu15,gu2016certikos,gu2018certikos,chen2016interrupts}, whose refinement proofs rely on
\textsc{CompCert}'s usermode-oriented memory model~\cite{leroy2008formal,leroy2009formally} which assumes
updates to one memory address are independent of updates to another ---
which is not true of page table updates.
 % which is not the case when a single page table write can make regions of memory inaccessible.
% \textsc{seL4}~\cite{Klein2009seL4,seL4TOCS,Sewell2013translation} is a formally verified L4 microkernel~\cite{Liedtke1995,Liedtke1996} 
% (and the first verified OS kernel to run on real-world hardware), verified with a mix of refinement proofs and program logic reasoning down to the assembly level.
% Because \textsc{seL4} is a microkernel, most VMM functionality actually lives in usermode and is unverified, and moreover, 
% their hardware model omits address translation entirely and the MMU entirely~\cite{Klein2009seL4,seL4TOCS}. 
% As a result, the limited page table management present in the microkernel treats page tables as idiosyncratic tree-maps as in
% \textsc{Verisoft}, despite actually running with address translation.
% This is partly mitigated by manually identifying some trusted invariants (e.g., that the address range designated for the kernel is appropriately mapped)
% and setting up the proof to ensure those invariants are maintained (i.e., as an extra proof obligation not required by their hardware model).

% \textsc{CertiKOS}~\cite{gu15,gu2016certikos,gu2018certikos,chen2016interrupts} is a microkernel intended for use as a hypervisor.
% The overall approach in that body of work is many layers of refinement proofs, using a
%  proliferation of layers with small differences to keep most individual refinements tractable. In keeping with precursor work 
% on the project from the same group~\cite{vaynberg2012compositional}, the purpose of some layers is to abstract away from 
% virtual memory (as early as possible).
% The papers on \textsc{CertiKOS} do not explicitly detail the VMM beyond highlighting its existence and referencing
% that it performs mapping operations for user code.
% The work is clear, however, that it fully trusts low-level assembly fragments such as the instruction sequence which actually
% switches address spaces, rather than verifying them.
% We can also infer other limitations from the project's pervasive reliance on a modified \textsc{CompCert},
% which fixes an abstract memory model aimed at \emph{user-mode} C programs~\cite{leroy2008formal,leroy2009formally}. 
% \textsc{CompCert}'s model of memory cannot handle non-local updates, so we can conclude \textsc{CertiKOS}'s proofs cannot
% address impact of virtual memory mapping changes on the kernel itself.
Other work on OS verification either never progressed far enough to address VMM verification 
(\textsc{Verisoft XT}~\cite{cohen2009vcc,cohen2010local,dahlweid2009vcc,cohen2013SOFSEM}), or uses memory-safe languages to enable safe co-habitation 
of a single address space by all processes (\textsc{Singularity}~\cite{Fahndrich2006language,Hunt2007singularity,Hunt2007sealing,Barnett2011specsharp}, \textsc{Verve}~\cite{Yang2010Verve},
and \textsc{Tock}~\cite{levy2017multiprogramming}). The latter approach ensures memory safety, but ignores other functional uses of virtual memory
hardware like swapping~\cite{Denning1970VM} or exploiting copy-on-write techniques for dynamic migration of virtual machines~\cite{clark2005live}.

\citet{kolanski08vstte} are the only other researchers to study VMM verification against a realistic hardware
model, where page table updates are performed through virtual memory accesses (later adding C-level type modeling~\cite{kolanski09tphols}).
As noted in Section \ref{sec:overly-restrictive}, Kolanski and Klein's virtual points-to is similar to that in Figure \ref{fig:strongvirtualpointsto},
with the attendant problems discussed earlier and lifted by our model. Their approach had modal elements,
but the work never addressed the issues of managing multiple address spaces, switching address spaces, or
invariants referring to other address spaces, which our explicitly modal approach and case studies do address.
They verified a fraction of an ARM version of our Figure \ref{fig:mapping_code}, which assumed they already had the virtual address
of the L1 entry (which they assumed existed), thereby skipping the physical-to-virtual conversion issues we treat in Section \ref{sec:p2v}.
Due to lack of fractional permissions in their formalism, they incidentally pick up other limitations orthogonal to their
foundational focus: by requiring
a virtual points-to have \emph{full} ownership of the page table walk memory, they limit their system to having only as many
virtual points-to assertions as there are entries in the top-level table (4096) because they cannot share access to entries.
We inherit fractional permission support form \iris, and use it extensively (the overly-restrictive
Figure \ref{fig:strongvirtualpointsto} is already an improvement in this way).

@@

% The other major distinction is that Kolanski and Klein have no explicit accounting for other address spaces.
% Their logic does not deal with change of address space, and has no way to assert that certain facts hold
% in another address space.
% They verify only one address space manipulation: mapping a single unmapped page into the current address space (in both papers).
% We verify this, as well as a change-of-address-space, which requires us to introduce assertions for talking
% about other address spaces (we must know, for example, that the precondition of the code after the change must be true
% in the \emph{other} address space), and to deal with the fact that the standard frame rule
% for separation logic is unsound in the presence of address space changes and address-space-contingent assertions.
% Our approach in this paper uses modalities to distinguish virtual-address-based assertions that hold only in specific 
% address spaces, making it possible to manipulate other address spaces, and equally critically, to \emph{change} address 
% spaces while reasoning about correctness.

Unlike our work, Kolanski and Klein prove very useful embedding theorems stating that code that does not modify page table 
entries can be verified in a VM-ignorant program logic, and that proofs in that logic can be embedded into the VM-aware logic 
(essentially by interpreting ``normal'' points-to relations as virtual points-to facts). While we have not proven such a result,
an analagous result {should} hold of our work: consider that the doubles for the \texttt{mov} instructions
that access memory behave just as one would expect for a VM-ignorant logic~\cite{Chlipala2013Bedrock}.
With our general approach to virtual points-to assertions being inspired by Kolanski and Klein, \emph{both}
 our approach and theirs could in principle be extended to account for pageable points-to assertions by adding additional 
disjunctions to an extended points-to definition; embedding ``regular'' separation logic into such a variant
is the appropriate next step to extend reasoning to usermode programs running with a kernel that may demand-page the program's
memory.

\paragraph{Translation Lookaside Buffers}
As noted in Section \ref{sec:backgroundonmachinemodel}, we do not formally model or reason about TLBs.
TLB flushes are required only when addresses are removed from a virtual address space, or when changing
virtual address spaces. Because this occurs in few places in the kernel (in some, only 3 locations), full verified
kernels including \textsc{seL4}~\cite{Klein2009seL4,seL4TOCS} and \textsc{CertiKOS}~\cite{gu15,gu2016certikos}
trust TLB management. Neither of the aforementioned systems has a hardware model including a TLB, so neither is able
to verify TLB management in any form --- they \emph{must} trust its operation.

The only place TLB code becomes particularly challenging is in multiprocessor kernels, where
not only must the running CPU flush its own TLB with special instructions, but it must send an inter-processor interrupt (IPI) to all
other cores to ensure they also flush the relevant ranges of their own TLBs.
Currently no formal hardware model exists with sufficient detail to reason about IPIs in full detail,
which would require extensions to model interactions with ACPI tables populated by firmware on desktop and server machines,
or flattened device trees for (semi-)embedded systems like single-board ARM and RISC-V machines, which tell the OS
important information about how to control other CPU cores;
plus the memory-mapped IO used to trigger those IPIs.
As a result, multicore extensions of \textsc{seL4}~\cite{von2013clustered} and \textsc{CertiKOS}~\cite{gu2016certikos} also trust this functionality.

% Fully grounded trust for this would require a formal model of how hardware populates ACPI tables and formal verification of ACPI
% parsing code on general-purpose desktop and server machines, or a formalized and verified
% parsing of flattened device trees (plus trust that a system was booted with a correct FDT) for (semi-)embedded systems like single-board
% ARM and RISC-V machines. Either route would also require detailed treatment of memory-mapped IO triggering
% interrupts on other CPUs. This is far beyond any formal model of computer hardware that exists today,
% and far out of scope for this paper.

% \citet{syeda2020formal,syeda2018program} are the only existing work to address formal verification of TLB management.
% They extend the work of \citet{kolanski08ssv,kolanski09tphols,kolanski09tphols}, and therefore inherit the limitations of that work discussed
% above. Their logic primarily tracks a set of mapped addresses and a set of
% cached addresses, and generally preserve a global invariant (in a global Hoare logic, not a separation logic)
% that the cached addresses are a subset of the mapped addresses. When this invariant is violated, the logic provides only
% a Hoare-style backwards assignment type rule with unconstrained assertion $P$, rather than providing structure in the logic
% for reasoning about how to restore the invariant. Most technical results in the paper focus on proving
% transparency lemmas, that code that \emph{does not} modify page tables (kernel code outside the VMM, user code)
% is unaffected by its existence. The one piece of kernel code verified is a sequence of 4 pseudo-instructions for changing address spaces
% which include \emph{only} the installation of a new page table root and flushing of the old address space from the TLB.

% Adapting Syeda and Klein's global reasoning principles to a separation logic is non-trivial, even though their approach
% does capture some intuition about TLB reasoning. And as noted above, in the absence of support for significant additional
% hardware functionality in the hardware model, this would be substantial work for only very limited gains in confidence compared
% to trusting single-core TLB managment.
% Thus we leave TLB management to our future work plans, where we believe it is best addressed in tandem with substantially
% richer hardware models than any currently existing.

The only other attempt to verify a change of address space was by \citet{syeda2018program,syeda2020formal}, whose focus was on TLB maintenance (which we
do not model). However, they verified only the 4 instructions to switch address spaces and update the TLB on an ARM processor, in isolation
(i.e., not the full context switch including changing stacks with address spaces).
The specification they proved for those instructions did not address program invariants that may be valid in one address space and not the other,
so is not flexible enough to extend directly to a full context switching primitive as in Figure \ref{fig:swtch},
in addition to inheriting other limitations from the work they extend, discussed above~\cite{kolanski08vstte,kolanski09tphols}.

The right eventual solution would be to combine our work
and theirs, which we leave to future work, as combining both the models and reasoning principles would be highly non-trivial.

\subsection{Program Logics with Modalities}
Modalities have long been a staple of program logics, at least as far back as Dijkstra's weakest precondition calculus~\cite{dijkstra1975guarded}
and Pratt's observation the Hoare triples
could be decomposed using the weakest-precondition modality of dynamic logic~\cite{pratt1976semantical},
in a form quite similar to what \iris uses today~\cite{jung2018iris}.
Variants of Nakano's later modality~\cite{nakano2000modality} have long been used to deal with step-indexing 
for impredicative and recursive features of logics and type systems~\cite{Appel2007,hobor2010theory,birkedal2011step,birkedal2013intensional,jung2018iris}.

As noted earlier, our other-space modality derives from hybrid logic~\cite{areces2001hybrid,blackburn1995hybrid,gargov1993modal,goranko1996hierarchies},
where modalities are indexed by \emph{nominals} which are names for specific individual states in a Kripke model.
Little prior work combines these ideas with program logics. \citet{brotherston2014parametric} show that traditional
nominals extends the expressive power of separation logic. \citet{gordon2019modal}
uses nominals to refer to states of other actors in an actor language.
In parallel with our work, \citet{wagner2024realistic} use a hybrid-logic-inspired modality
to abstract reference-counting specifics from specifications of a low-level application binary interface (ABI)
--- their $@_l(P)$ indicates that $l$ is the location of a reference count for resources satisfying $P$.
% As noted throughout the paper, the inspiration for our other-space modality comes from hybrid logic~\cite{areces2001hybrid,blackburn1995hybrid,gargov1993modal,goranko1996hierarchies},
% where modalities are indexed by \emph{nominals} which are names for specific individual states in a Kripke model.
% We are aware of only two prior works combining hybrid logics with program logics specifically. 
% Brotherston and Villard~\cite{brotherston2014parametric} demonstrated that may properties true of various 
% separation logics are not definable in boolean \BI (\BBI), and showed that a hybrid extension \HyBBI allows
% most such properties to be defined (e.g., the fact that separating conjunction is cancellative is unprovable 
% in boolean \BI, but provable in \HyBBI). There, nominals named resources 
% (roughly, but not exactly, heap fragments). 
% Gordon~\cite{gordon2019modal} described a use of hybrid logic in the verification of actor programs, 
% where nominals named the local state of individual actors (with such assertions stabilized with a 
% rely/guarantee approach). 
Beyond these, there is limited work on the interaction of hybrid logic with general substructural logics, in restricted forms
that do not affect expressivity~\cite{despeyroux2014hybrid,chaudhuri2019hybrid}.\looseness=-1
% Primarily there is a line of work on hybrid linear logic (\HyLL)~\cite{despeyroux2014hybrid}, 
% originally used as a way to more conveniently express aspects of transition systems in linear logic. 
% However, \HyLL's proof rules offer no non-trivial interactions with multiplicative connectives 
% (every \HyLL proof can in fact be embedded into regular linear logic~\cite{chaudhuri2019hybrid}, 
% unlike Brotherston and Villard's \HyBBI, which demonstrably increases expressive power over its base \BBI.
